---
title: "RandomSampling"
author: "SLW"
date: "March 20, 2017"
output: 
  html_document: 
    keep_md: yes
---

Load Packages
```{r LOAD PACKAGES, include=FALSE,results='hide'}
#source("https://bioconductor.org/biocLite.R")
#biocLite('dplyr')
  library(Rcpp)
  library(methylumi)
  library(lumi)
  library(minfi)
  library(gplots)
  library(reshape2)
  library(reshape)
  library(ggplot2)
  library(RColorBrewer)
  library(gridExtra)
  library(grid)
  library(MASS)
  library(limma)
  library(plyr)
  library(dplyr)
```

```{r Load Data}
memory.limit(10000000000000)
setwd('Z:/ROBLAB1 coredata-databases/1 Samantha DATA Folder/PROJECTS/DNAmProfiling_PE_IUGR_2017')

##Read in phenotype data
des<-read.csv('CoX_Samples_2017.csv',header=T)
for (i in 1:nrow(des)){
  des$Row[i]<-paste(substr(des[i,"Sentrix_Position"], start=1, stop=3))
}
des$Row<- as.factor(des$Row)
str(des)
des$Sentrix_ID<-as.factor(des$Sentrix_ID)
des$GA<-as.numeric(des$GA)

##As IUGR is not fully represented in all 6 row positions, I group rows into levels 1,2,and 3 on the microarray
des$row_grouped<-revalue(des$Row,c("R01"="1","R02"="1","R03"="2","R04"="2","R05"="3",
                                   "R06"="3"))
des$row_grouped<-as.numeric(des$row_grouped)
rownames(des)<-des$ParticipantID

##Loading in DNAm Data- DNAm measured on the Illumina 450K array
load('CoxCohort.fnorm_Jan2016.RData')
Data<-exprs(PROJECT.fun)
dim(Data)

Data<-as.data.frame(Data)
Data$COX_9801<-Data$CPX_9801
Data$CPX_9801<-NULL
Data$COX_11387<-Data$COX_11378
Data$COX_11378<-NULL
Data<-as.matrix(Data)

all(colnames(Data)==rownames(des))##FALSE
Data<-Data[,rownames(des)]
all(colnames(Data)==rownames(des))##TRUE
```

Question: Of the 1703 EOPE differentially methylated sites in the discvoery cohort, 599 (35.1%) validated in the validation cohort. Is this more than we would expect by chance?

39 probes met both FDR and Delta beta thresholds in both cohorts, is this more than we would expect by chance?

To look into this I will sample 1703 random CpG sites from the validation cohort, run the linear model, and record how many validate, 1000 times. Then I will compare this average to the actual number of hits that validated to determine if it is more than we would expect by chance.

```{r}
##Take a rondom 1703 probes from the Validation cohort data, run a linear model correcting for fetal sex
random<-function(A){
  B<-A[sample(440093,replace=TRUE),]
  lapply(1:nrow(B), function(CpG){
  pheno<-des
  pheno$Mval<-B[CpG,]
  model_hits<-lm(Mval~GRP+Sex, data=pheno)
  coef(summary(model_hits))[2,4]}) ##pulls row 2, column 4 corresponds to EOPEvsPreT (Row 2), p-values (Column 4)
}

##Perform 1000 iterations of random sampling and linear modelling
p.vals<-as.vector(replicate(1000,random(Data))) ##This will run the random function 1000 times

##For each column in p.vals, how many meet nominal p<0.05
Sig<-as.vector(colSums(p.vals<0.000001))

##Number of hits that were actually validated in the data
PH<-as.numeric(38)

##Compare the number of "validated" hits in the random sampling the the actual number of validated hits. Are they significantly different.
#Source code- Adapted from Rachel Edgar
perm_num<-1000 ##Number of iterations
sum(Sig>PH) + 1/(perm_num + 1)##p=0.000999001

##Yes the number of validated hits we obtain in the analysis is singificantly more than we expect by chance. 
```

